Race Science in Mainstream Psychology, 1960-2010

### 2024-06-20 paper/paper.qmd
re-snapshot renv bc compiler errors

---
title: "Race Science in Mainstream Psychology, 1960-2010"
authors:
  - name: "Daniel J. Hicks"
    email: "dhicks4@ucmerced.edu"
    affiliations: 
      - id: a
        name: "University of California, Merced"
        city: "Merced"
        country: "CA, USA"
    orcid: "0000-0001-7945-4416"
    corresponding: true
  - name: "Emilio J. C. Lobato"
    affiliations:
      - ref: a
    orcid: "*[todo]*"

@@

figure sizes: 
Small: approximately 9 cm x 6 cm
Medium: approximately 11 cm x 11 cm
Large: approximately 18 cm x 22 cm

@@

*[TODO: 
- NHB reviews todo: <https://docs.google.com/document/d/1xzG95pZOZgZQFfSk6yYMFhrnUQ8AWk9U1O5ngm3n_uo/edit>
- copy updated plots
- figure sizes
- align abstract w/ intro
- figures and tables to end
]*

@@

We stress that race science discourse includes methodological and empirical critiques of race science.  We argue that such technical critiques — even when they are intended to undermine race science — can instead legitimize it, by creating an impression among the broader public that race science is a matter of ordinary, reasonable, mainstream scientific debate.  For example, the Flynn effect (the finding that IQ scores have increased over time) poses a serious empirical challenge to hereditarian claims of racial differences in intelligence, and has been extensively debated in the mainstream scientific literature.  Most members of the general public lack the time and skills needed to assess these debates for themselves.  So, we suggest, they are quite likely to use a "middle ground" heuristic to understand these debates as ordinary, reasonable disagreement among mainstream scientists.  Some experts say one thing, other experts say something else, and so "the truth is somewhere in the middle."  Tobacco and other industries have exploited this heuristic to "manufacture doubt" about the hazards of their products *[cites]*.  In this way technical critiques of race science based on the Flynn effect could legitimize race science.  So, while empirical research supporting the Flynn effect is usually not itself race science, it can still be race science discourse.

@@

In light of the scientific racist origins of Pioneer and MQ, and the significant attention MQ has received in recent historiography, we hypothesized that MQ might have served as an "incubator" for race science discourse, with ideas originating in MQ before appearing within the scientific mainstream.  Establishing venues for favorable research is another key element of the "tobacco strategy" for manufacturing doubt *[FP]*.  Specifically, we hypothesized that bibliometric and text mining analyses would show race science discourse appearing first in MQ, and subsequently appearing in mainstream publications.  This hypothesis was not supported.  Instead, mainstream psychology journals provided a venue for a distinct form of race science discourse, based on intelligence research and originating outside of MQ.

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results.](img/wos_results.png){#fig-wos width="4.76in" height="2.6in" fig-env="figure*"}

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/focal_topics.png){#fig-focal width="4.76in" height="3.57in" fig-env="figure*"}

@@

![Count of documents in the corpus, by journal and year.](img/02_count.png){#fig-counts width="11cm" height="11cm" fig-env="figure*"}

@@

![Silge plots (top 15 terms) for each topic and model, large vocabulary.](supplements/S2-lg-silge.pdf){#fig-silge-lg fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, medium vocabulary.](supplements/S3-md-silge.pdf){#fig-silge-md fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, small vocabulary.](supplements/S4-sm-silge.pdf){#fig-silge-sm fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, large vocabulary.](supplements/S5_grid_lg.pdf){#fig-grid-lg fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, medium vocabulary.](supplements/S6_grid_md.pdf){#fig-grid-md fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, small vocabulary.](supplements/S7_grid_sm.pdf){#fig-grid-sm fig-env="figure*"}

>>>>

---
title: "Race Science in Mainstream Psychology, 1960-2010"
authors:
  - name: "Daniel J. Hicks"
    email: "dhicks4@ucmerced.edu"
    affiliations: 
      - id: a
        name: "University of California, Merced"
        city: "Merced"
        country: "CA, USA"
    orcid: "0000-0001-7945-4416"
    corresponding: true
  - name: "Emilio J. C. Lobato"
    affiliations:
      - ref: a
    orcid: "0000-0002-3066-2932"

@@

figure sizes: 
Small: approximately 9 cm x 6 cm (3.5 in x 2.4 in)
Medium: approximately 11 cm x 11 cm (4.3 in x 4.3 in)
Large: approximately 18 cm x 22 cm (7.1 in x 8.7 in)

@@

*[TODO: 
- figure out renv::restore()
    - re-make
    - copy updated plots
]*

@@

We stress that race science discourse includes methodological and empirical critiques of race science.  We argue that such technical critiques — even when they are intended to undermine race science — can instead legitimize it, by creating an impression among the broader public that race science is a matter of ordinary, reasonable, mainstream scientific debate.  For example, the Flynn effect (the finding that IQ scores have increased over time) poses a serious empirical challenge to hereditarian claims of racial differences in intelligence, and has been extensively debated in the mainstream scientific literature.  Most members of the general public lack the time and skills needed to assess these debates for themselves.  So, we suggest, they are quite likely to use a "middle ground" heuristic to understand these debates as ordinary, reasonable disagreement among mainstream scientists.  Some experts say one thing, other experts say something else, and so "the truth is somewhere in the middle."  Tobacco and other industries have exploited this heuristic to "manufacture doubt" about the hazards of their products [@FernandezPintoKnowBetterNot2017].  In this way technical critiques of race science based on the Flynn effect could legitimize race science.  So, while empirical research supporting the Flynn effect is usually not itself race science, it can still be race science discourse.

@@

In light of the scientific racist origins of Pioneer and MQ, and the significant attention MQ has received in recent historiography, we hypothesized that MQ might have served as an "incubator" for race science discourse, with ideas originating in MQ before appearing within the scientific mainstream.  Establishing venues for favorable research is another key element of the "tobacco strategy" for manufacturing doubt [@FernandezPintoKnowBetterNot2017].  Specifically, we hypothesized that bibliometric and text mining analyses would show race science discourse appearing first in MQ, and subsequently appearing in mainstream publications.  This hypothesis was not supported.  Instead, mainstream psychology journals provided a venue for a distinct form of race science discourse, based on intelligence research and originating outside of MQ.

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results.](img/pf_wos_results.png){#fig-wos width="7.1in" height="3.5in" fig-env="figure*"}

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/08_focal_topics.png){#fig-focal width="7.1in" height="5.2in" fig-env="figure*"}

@@

![Count of documents in the corpus, by journal and year.](img/02_count.png){#fig-counts width="3.5in" height="2.4in" fig-env="figure*"}

@@

![Silge plots (top 15 terms) for each topic and model, large vocabulary.](img/06-lg-silge.pdf){#fig-silge-lg fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, medium vocabulary.](img/06-md-silge.pdf){#fig-silge-md fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, small vocabulary.](img/06-sm-silge.pdf){#fig-silge-sm fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, large vocabulary.](img/08_grid_lg.pdf){#fig-grid-lg fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, medium vocabulary.](img/08_grid_md.pdf){#fig-grid-md fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, small vocabulary.](img/08_grid_sm.pdf){#fig-grid-sm fig-env="figure*"}
